---
title: "Accept-Reject-digression-tldr"
description: |
  Accept-Reject intuitive description.
draft: false
categories: [MCMC, Accept-Reject]
author:
  - name: Robert Settlage
    orcid: 0000-0002-1354-7609
date: 2022-06-04
output:
  distill::distill_article:
    self_contained: false
    highlight: default
    toc: true
    float: true
includes:
  \usepackage{amsmath}
editor_options: 
  markdown: 
    wrap: 72
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
library(knitr)
library(kableExtra)
library(ggpubr)
library(ggplot2)
library(ggExtra)
```


Suppose we have an area $\Omega'$ completely enclosed by another area $\Omega$ as seen below.  We know how to sample from the larger area $\Omega$ but we really want to sample from $\Omega'$.  

```{r}
plot(1, 1, col = "white", xlab = "X", ylab = "Y", xlim=c(0.5,2), ylim=c(0,1.1))  
polygon(x = c(0.7, 1.7, 1.1, 0.8),
        y = c(0.6, 1, 0.7, .1),
        col = "#1b98e0") 
```

We can do this via the Accept-Reject algorithm as follows:

Let  
$g(x) = \frac{\mathbb{1}_\Omega (x)}{A(\Omega)}$ where we have an indicator for in/out of $\Omega$ and A is the area defined by $\Omega$.  

Similarly, we define our target by:  
$f(x) = \frac{\mathbb{1}_\Omega' (x)}{A(\Omega')}$

The AR algorithm is then:

1. draw $u\sim unif(0,1)$
2. draw $x \sim g(\cdot)$
3. accept if $u \le \frac{f(x)}{Mg(x)} = \frac{A(\Omega)}{A(\Omega')}\frac{\mathbb{1}_\Omega' (x)}{\mathbb{1}_\Omega (x)}\frac{1}{M}$.  We should pick $M=\frac{A(\Omega)}{A(\Omega')}$.


## Example

Darts to estimate $\pi$.

Consider a circle circumscribed within the bounds of the square, both centered on (x,y)=(0,0).  Let the radius of the circle be 1 such that if we focus on the first quadrant, we are looking at a unit square.  Within this quadrant, we can form the ratio of the areas as:

$$
\begin{equation}
\frac{\tfrac{1}{4}A_{circle}}{A_{square}} = \frac{\pi r^2}{4r^2} = \frac{\pi}{4}
\end{equation}
$$
So, if we could measure the area of the circle vs square in this quadrant, we know the ratio will be $\frac{\pi}{4}$.  So, how can we determine the areas?  How about we sample from the unit square and perform accept-reject if the points samples fall within the circle?

1. draw $u_1 \sim unif(0,1), u_2\sim unif(0,1)$  
2. accept if $\sqrt{u_1^2 + u_2^2} \le 1$  
3. $\pi = 4*\frac{accept}{draws}$

```{r echo=TRUE, include=TRUE,fig.height=6, fig.width=6, fig.align = 'center'}
set.seed(189437)
N <- 1e6
accept <- rep(0,N)
u1_draw <- runif(N)
u2_draw <- runif(N)
for(i in 1:N){
  if(sqrt(u1_draw[i]^2+u2_draw[i]^2)<1){
    accept[i] <- 1
  }
}
simpi <- 4*sum(accept)/N
plot(u1_draw, u2_draw, col=accept+1,pch=20)
text(0.4,0.4,paste0(N," draws to estimate pi to ",simpi))
```


