---
title: "tl;dr Inverse CDF"
description: |
  Monte Carlo using the Inverse CDF.
draft: false
categories: [MCMC, Inverse CDF]
author:
  - name: Robert Settlage
    orcid: 0000-0002-1354-7609
date: 2022-06-11
output:
  distill::distill_article:
    self_contained: false
    highlight: default
    toc: true
    float: true
includes:
  \usepackage{amsmath}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
library(knitr)
library(kableExtra)
library(ggpubr)
library(ggplot2)
library(ggExtra)
```

Summary of the Inverse CDF method for generating draws from a distribution given draws from a random uniform. 

## Proposition

The random variable $Y=F^{-1}(u)$, where $F$ is a distribution function and $u\sim unif(0,1)$, will have distribution $F$.


## Proof

I like the proof that this is true.  It is truly simple:

Let $G_Y(y)$ be a distribution function for Y.  Then:

$$
\begin{eqnarray}
\tag{1}
G_Y(\alpha) &=& Pr(y \le \alpha) \\
  &=& Pr(F^{-1}(u) \le \alpha) \\
  &=& Pr(u \le F(\alpha)) \\
  &=& F(\alpha)
\end{eqnarray}
$$

## Example 1

Generate samples from $x\sim exp(\lambda)$ using draws from $u\sim unif(0,1)$.


+ Step 1: Find CDF

$$
\begin{equation}
\tag{2}
F(\alpha) = \int_0^{\alpha} \lambda e^{-\lambda x} dx = e^{-\lambda x} \mid_0^{\alpha} = 1-e^{\lambda\alpha}
\end{equation}
$$

+ Step 2: Find invserse CDF

let $g=1-e^{\lambda\alpha}$

$$
\begin{eqnarray}
\tag{3}
g &=& 1-e^{\lambda\alpha} \\
e^{-\lambda \alpha}  &=& 1-g \\
\alpha &=& -\frac{1}{\lambda}log(1-g) \\
F^{-1}(\alpha) &=& -\frac{1}{\lambda}log(1-\alpha)
\end{eqnarray}
$$

+ Step 3: Generate Y

draw $u_i \sim unif(0,1)$, now, $Y=-\frac{1}{\lambda}log(1-u) \sim exp(\lambda)$

## Example 2

What if the desired distribution is discrete?  For instance, $bern(p)$.

$$
\begin{equation}
\tag{4}
x \sim 
\begin{cases}
1 \text{ with probability } p \\
0 \text{ with probability } 1-p
\end{cases}
\end{equation}
$$
Graphically, this looks like:

```{r bern_graph, echo=TRUE, include=TRUE}
#for graphing, choose arbitrary p=0.6
p=0.75

# Create empty example plot
plot(0, 0, col = "white", xlab = "y", ylab = "u", xlim=c(0,2), ylim=c(0,1), xaxt="n", yaxt="n")
axis(1, at=c(0.5,1.5),labels=c("0","1"), col.axis="red", las=1)
axis(2, at=c(1-p,p),labels=c("1-p","p"), col.axis="red", las=2)
# Draw one line
segments(x0 = 0, y0 = 1-p, x1 = 1-0.03, y1 = 1-p, col = "red",lwd=2)
segments(x0 = 1, y0 = p, x1 = 2, y1 = p, col = "blue",lwd=2)
segments(x0 = 1, y0 = 1-p+0.06, x1 = 1, y1 = p, col = "black",lwd=2, lty="dotted")
points(1,1-p,pch=1,col="red",cex=2)
points(1,p,pch=20,col="blue",cex=2)
text(0.5,0.3,"1-p")
text(1.5,0.8,"p")
```

This is relatively easy:

+ Step 1: 
draw $u_i \sim unif(0,1)$

+ Step 2:

$$
\begin{equation}
\tag{4}
y = 
\begin{cases}
0 \text{ if } u \lt 1-p \\
1 \text{ otherwise }
\end{cases}
\end{equation}
$$

## Example 3

An approximate CDF can also be generated in the same way.  Think of this as a discretized continuous random variable.

Suppose we want to generate $y\sim N(0,1)$.

$$
\begin{equation}
\tag{4}
f(y_i) = 
\begin{cases}
f(y_0) \text{ if } f(y_0) \le u \lt f(y_1) \\
f(y_1) \text{ if } f(y_1) \le u \lt f(y_2) \\
f(y_2) \text{ if } f(y_2) \le u \lt f(y_3) \\
\vdots
\end{cases}
\end{equation}
$$

Graphically, this starts to look like:

```{r desc_norm_graph, echo=TRUE, include=TRUE}
# function for arbitrary pdf
pdf <- function(x){
  y <- (1/(2*pi))*exp(-0.5*x^2)
}
par(mfcol=c(1,2))

## PDF
plot(0, 0, col = "white", xlab = "", ylab = "", xlim=c(-4,4), ylim=c(0,0.2), main="Discretized PDF")
curve(pdf, from=-4, to=4, xlab="x", ylab="y", add=TRUE)
segments(x0 = -2, y0 = 0, x1 = -2, y1 = pdf(-2), col = "red",lwd=2)
segments(x0 = -1.75, y0 = 0, x1 = -1.75, y1 = pdf(-1.75), col = "red",lwd=2)
segments(x0 = -1.5, y0 = 0, x1 = -1.5, y1 = pdf(-1.5), col = "red",lwd=2)
segments(x0 = -1.25, y0 = 0, x1 = -1.25, y1 = pdf(-1.25), col = "red",lwd=2)
segments(x0 = -1, y0 = 0, x1 = -1, y1 = pdf(-1), col = "red",lwd=2)

# CDF
plot(0, 0, col = "white", xlab = "", ylab = "u", xlim=c(-2,0), ylim=c(0,0.2),main="Inverse CDF step function")
segments(x0 = -2, y0 = pdf(-2), x1 = -1.75, y1 = pdf(-2), col = "red",lwd=2)
segments(x0 = -1.75, y0 = pdf(-1.75), x1 = -1.5, y1 = pdf(-1.75), col = "red",lwd=2)
segments(x0 = -1.5, y0 = pdf(-1.5), x1 = -1.25, y1 = pdf(-1.5), col = "red",lwd=2)
segments(x0 = -1.25, y0 = pdf(-1.25), x1 = -1, y1 = pdf(-1.25), col = "red",lwd=2)
segments(x0 = -1, y0 = pdf(-1), x1 = -0.75, y1 = pdf(-1), col = "red",lwd=2)
points(y = pdf(-2), x = -1.75,pch=1,col="red")
points(y = pdf(-1.75), x = -1.5,pch=1,col="red")
points(y = pdf(-1.5), x = -1.25,pch=1,col="red")
points(y = pdf(-1.25), x = -1,pch=1,col="red")
points(y = pdf(-1), x = -0.75,pch=1,col="red")

```

To get the final points, they must be normalized by the total of the draws:

$$
\begin{equation}
f(y_i) = \frac{e^{-\frac{1}{2}y_i^2}}{\sum_{j=0}^N e^{-\frac{1}{2}y_j^2}} 
\end{equation}
$$






